---
title: "Chicago Crime Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: cosmo
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
library(plotly)
library(leaflet)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(DT)

# Load pre-processed data
crime_data <- readRDS("crime_data_clean.rds")
coords_clusters <- readRDS("coords_clusters.rds")
kmeans_model <- readRDS("kmeans_model.rds")
rf_model <- readRDS("rf_model.rds")
```

# Overview

## Row {data-height=120}

### Total Crimes

```{r}
valueBox(
  value = format(nrow(crime_data), big.mark = ","),
  icon = "fa-gavel",
  color = "danger"
)
```

### Arrest Rate

```{r}
arrest_rate <- round(mean(crime_data$Arrest, na.rm = TRUE) * 100, 1)
valueBox(
  value = paste0(arrest_rate, "%"),
  icon = "fa-handcuffs",
  color = ifelse(arrest_rate > 30, "success", "warning")
)
```

### Most Common Crime

```{r}
top_crime <- names(sort(table(crime_data$PrimaryType), decreasing = TRUE))[1]
valueBox(
  value = top_crime,
  icon = "fa-exclamation-triangle",
  color = "primary"
)
```

## Row {data-height=880}

### Top 10 Crime Types

```{r}
top10 <- crime_data %>%
  count(PrimaryType, sort = TRUE) %>%
  head(10) %>%
  mutate(PrimaryType = reorder(PrimaryType, n))

p1 <- plot_ly(
  top10,
  x = ~n,
  y = ~PrimaryType,
  type = "bar",
  orientation = "h",
  marker = list(
    color = ~n,
    colorscale = list(c(0, "#440154"), c(0.5, "#21918c"), c(1, "#fde725")),
    line = list(color = "rgba(0,0,0,0.3)", width = 0.5)
  ),
  text = ~paste0(format(n, big.mark = ","), " crimes"),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Top 10 Crime Types", font = list(size = 16)),
    xaxis = list(title = "Number of Crimes", tickformat = ","),
    yaxis = list(title = ""),
    margin = list(l = 150)
  )
p1
```

### Crimes by Hour of Day

```{r}
hourly <- crime_data %>%
  count(Hour) %>%
  mutate(Hour = as.integer(Hour))

p2 <- plot_ly(
  hourly,
  x = ~Hour,
  y = ~n,
  type = "scatter",
  mode = "lines+markers",
  line = list(color = "#e63946", width = 3, shape = "spline"),
  marker = list(size = 7, color = "#e63946"),
  text = ~paste0("Hour: ", Hour, ":00\nCrimes: ", format(n, big.mark = ",")),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Crime Distribution by Hour", font = list(size = 16)),
    xaxis = list(title = "Hour of Day", dtick = 2, range = c(-0.5, 23.5)),
    yaxis = list(title = "Number of Crimes", tickformat = ",")
  )
p2
```

# Trends

## Row {data-height=500}

### Crimes by Month

```{r}
monthly <- crime_data %>%
  count(Month) %>%
  mutate(MonthName = month.abb[Month])

month_order <- month.abb
monthly$MonthName <- factor(monthly$MonthName, levels = month_order)

p3 <- plot_ly(
  monthly,
  x = ~MonthName,
  y = ~n,
  type = "bar",
  marker = list(
    color = ~n,
    colorscale = "Portland",
    line = list(color = "rgba(0,0,0,0.2)", width = 0.5)
  ),
  text = ~paste0(MonthName, ": ", format(n, big.mark = ",")),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Crimes by Month", font = list(size = 16)),
    xaxis = list(title = "Month"),
    yaxis = list(title = "Number of Crimes", tickformat = ",")
  )
p3
```

### Crimes by Season

```{r}
seasonal <- crime_data %>%
  count(Season) %>%
  mutate(Season = factor(Season, levels = c("Spring", "Summer", "Fall", "Winter")))

season_colors <- c("Spring" = "#2ecc71", "Summer" = "#e74c3c", "Fall" = "#e67e22", "Winter" = "#3498db")

p4 <- plot_ly(
  seasonal,
  x = ~Season,
  y = ~n,
  type = "bar",
  marker = list(color = season_colors[levels(seasonal$Season)]),
  text = ~paste0(Season, ": ", format(n, big.mark = ",")),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Crimes by Season", font = list(size = 16)),
    xaxis = list(title = "Season"),
    yaxis = list(title = "Number of Crimes", tickformat = ",")
  )
p4
```

## Row {data-height=500}

### Crimes by Year

```{r}
yearly <- crime_data %>%
  count(Year)

p5 <- plot_ly(
  yearly,
  x = ~Year,
  y = ~n,
  type = "scatter",
  mode = "lines+markers",
  line = list(color = "#6c5ce7", width = 3),
  marker = list(size = 6, color = "#6c5ce7"),
  text = ~paste0("Year: ", Year, "\nCrimes: ", format(n, big.mark = ",")),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Crime Trends by Year", font = list(size = 16)),
    xaxis = list(title = "Year", dtick = 2),
    yaxis = list(title = "Number of Crimes", tickformat = ",")
  )
p5
```

### Crime Type vs Time of Day Heatmap

```{r}
# Get top 10 crime types for readability
top_types <- crime_data %>%
  count(PrimaryType, sort = TRUE) %>%
  head(10) %>%
  pull(PrimaryType)

heatmap_data <- crime_data %>%
  filter(PrimaryType %in% top_types) %>%
  count(PrimaryType, TimeOfDay) %>%
  tidyr::pivot_wider(names_from = TimeOfDay, values_from = n, values_fill = 0)

time_periods <- intersect(c("Morning", "Afternoon", "Evening", "Night"), colnames(heatmap_data))

heat_matrix <- as.matrix(heatmap_data[, time_periods])
rownames(heat_matrix) <- heatmap_data$PrimaryType

p6 <- plot_ly(
  x = time_periods,
  y = heatmap_data$PrimaryType,
  z = heat_matrix,
  type = "heatmap",
  colorscale = "YlOrRd",
  text = format(heat_matrix, big.mark = ","),
  hovertemplate = "Crime: %{y}<br>Time: %{x}<br>Count: %{text}<extra></extra>"
) %>%
  layout(
    title = list(text = "Crime Type vs Time of Day", font = list(size = 16)),
    xaxis = list(title = "Time of Day"),
    yaxis = list(title = "", tickfont = list(size = 10)),
    margin = list(l = 150)
  )
p6
```

# Hotspot Map

## Row

### Crime Hotspot Clusters (K-Means, K=5)

```{r}
set.seed(42)
map_sample <- coords_clusters %>%
  sample_n(min(10000, nrow(coords_clusters)))

cluster_colors <- c("#e74c3c", "#3498db", "#2ecc71", "#f39c12", "#9b59b6")
cluster_pal <- colorFactor(cluster_colors, domain = 1:5)

leaflet(map_sample) %>%
  addProviderTiles(providers$CartoDB.DarkMatter) %>%
  addCircleMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    radius = 3,
    color = ~cluster_pal(Cluster),
    fillOpacity = 0.6,
    stroke = FALSE,
    popup = ~paste0(
      "<strong>Crime Type:</strong> ", PrimaryType, "<br>",
      "<strong>Cluster:</strong> ", Cluster, "<br>",
      "<strong>Location:</strong> ", round(Latitude, 4), ", ", round(Longitude, 4)
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = cluster_pal,
    values = 1:5,
    title = "Cluster Zone",
    opacity = 0.8
  ) %>%
  setView(lng = -87.68, lat = 41.85, zoom = 11)
```

# ML Results

## Row {data-height=500}

### Random Forest: Confusion Matrix

```{r}
conf_matrix <- rf_model$confusion
# Remove class.error column for display
conf_display <- conf_matrix[, 1:(ncol(conf_matrix) - 1)]

# Create a styled table
knitr::kable(
  conf_matrix,
  caption = "Random Forest Confusion Matrix (Arrest Prediction)",
  format = "html",
  align = "c"
) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center",
    font_size = 14
  )
```

### Random Forest: Feature Importance

```{r}
importance_df <- data.frame(
  Feature = rownames(rf_model$importance),
  Importance = rf_model$importance[, "MeanDecreaseGini"]
) %>%
  arrange(desc(Importance)) %>%
  mutate(Feature = reorder(Feature, Importance))

p7 <- plot_ly(
  importance_df,
  x = ~Importance,
  y = ~Feature,
  type = "bar",
  orientation = "h",
  marker = list(
    color = ~Importance,
    colorscale = list(c(0, "#667eea"), c(1, "#764ba2")),
    line = list(color = "rgba(0,0,0,0.2)", width = 0.5)
  ),
  text = ~paste0(Feature, ": ", round(Importance, 1)),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Feature Importance (Mean Decrease Gini)", font = list(size = 16)),
    xaxis = list(title = "Mean Decrease Gini"),
    yaxis = list(title = ""),
    margin = list(l = 120)
  )
p7
```

## Row {data-height=500}

### Linear Regression Note

```{r}
htmltools::div(
  style = "padding: 30px; font-size: 16px; line-height: 1.8;",
  htmltools::h3("Linear Regression Analysis", style = "color: #2c3e50;"),
  htmltools::hr(),
  htmltools::p(
    htmltools::strong("R² = 0.083"), " — The linear regression model explains only ~8.3% of the variance in arrest outcomes."
  ),
  htmltools::h4("Why is R² so low?", style = "color: #e74c3c; margin-top: 20px;"),
  htmltools::tags$ul(
    style = "font-size: 15px;",
    htmltools::tags$li("Crime arrest outcomes are inherently non-linear and depend on complex interactions between features."),
    htmltools::tags$li("Linear regression assumes a linear relationship between predictors and the response, which does not hold well for binary classification tasks like arrest prediction."),
    htmltools::tags$li("Many important factors influencing arrests (e.g., police response time, evidence availability, witness cooperation) are not captured in the dataset."),
    htmltools::tags$li("The Random Forest model (non-linear) significantly outperforms linear regression, confirming non-linear relationships in the data.")
  ),
  htmltools::h4("Key Takeaway", style = "color: #27ae60; margin-top: 20px;"),
  htmltools::p("Non-linear models like Random Forest are far more suitable for this type of classification problem. The low R² of linear regression validates the choice of ensemble methods for this use case.")
)
```

### K-Means: Cluster Size Distribution

```{r}
cluster_sizes <- data.frame(
  Cluster = paste("Cluster", 1:5),
  Size = kmeans_model$size
)

p8 <- plot_ly(
  cluster_sizes,
  labels = ~Cluster,
  values = ~Size,
  type = "pie",
  textposition = "inside",
  textinfo = "label+percent",
  marker = list(
    colors = c("#e74c3c", "#3498db", "#2ecc71", "#f39c12", "#9b59b6"),
    line = list(color = "#FFFFFF", width = 2)
  ),
  hovertemplate = "%{label}<br>Size: %{value:,}<br>Percentage: %{percent}<extra></extra>"
) %>%
  layout(
    title = list(text = "K-Means Cluster Size Distribution (K=5)", font = list(size = 16)),
    showlegend = TRUE
  )
p8
```
